Can Auxiliary Indicators Improve COVID-19 Forecasts and Hotspot Prediction?

### 2021-06-14 forecast/paper/paper.tex
Edits through results summary

There is another, more subtle issue in evaluating predictive utility that 
deserves explicit mention, as it will play a key role in our analysis.
Signals computed from surveillance streams will often be subject to  
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and/or processed late. Correlations between such a signal and case
rates calculated ``after the fact'' (computed retrospectively, using the
finalized values of this signal) will not deliver an honest answer to the    
question of whether this signal would have been useful in real time. Instead,
we build predictive models using only the data that would have been available
\textit{as of} the prediction date, and compare the ensuing predictions in terms 
of accuracy. To do so, we leverage Delphi's \texttt{evalcast} R package 
\cite{EvalcastR}, which plugs into the COVIDcast API's data versioning system,  
and facilitates honest backtesting.

% Furthermore, adopting this practically-oriented forecasting
% perspective to indicator evaluation requires one to confront
% fundamental limitations inherent to public health surveillance
% systems.  In particular, many reporting systems are plagued by
% latency, with delays in reporting that can sometimes span
% multiple weeks.  Latency manifests itself in different ways in different data sources.  Some data sources may only make a signal available one week after the date it describes; other
% data sources may provide immediate imperfect initial estimates, which are
% gradually ``backfilled'' over a period of days, weeks, or months (in which previous estimates are updated as new
% data trickle in).  And few of
% these latency patterns are stable or predictable, with holidays,
% extreme weather events, and changes in the reporting process all
% affecting the dynamics of the delay process.  To fully and accurately
% understand an indicator's usefulness for forecasting requires
% accounting for the unfortunate reality of data latency.

% Examining the importance of additional features for prediction is a core question
% in inferential statistics, and there are many ways to examine this question
% under a variety of assumptions. For time series, work in
% this area can be traced at least to Granger causality~\cite{GrangerCausality} and
% related multivariate extensions~\cite{GewekeTimeSeries} under the condition 
% that the series are linearly related autoregressions among other assumptions.
% These techniques are especially popular in econometrics and neuroscience~\cite{mccracken2007asymptotics,NeuroscienceGranger}.

% We choose to take a
% predictive angle, which is simple but also fairly robust, compared to various
% other classical inferential approaches. In fact, drawing rigorous inference based on
% predictions, without (or with lean) assumptions, is an active field of
% research~\cite{dai2021significance,fryer2021shapley,zhang2020floodgate,williamson2020unified}.
% Nonetheless, the predictive approach directly examines the target of interest: do
% these signals help to produce forecasts? Furthermore, directly comparing
% forecast errors without strong assumptions through tests of predictive accuracy
% are common in epidemiology, econometrics, and other
% fields~\cite{DieboldMarianoTest,DieboldRetrospective}.

% - then make the broad point that "feature importance" is a core question in
% statistics, and there are many ways to look at it and lots of literature (cite
% a bunch fo stuff).  We choose to take a predictive angle, which is simple but
% also fairly robust, compared to various other classical inferential approaches
% that make lots of assumptions.  In fact, drawing rigorous inference based on
% predictions, without (or with lean) assumptions, is an active field of
% research.  Cite several papers (Jing's conformal paper that describes LOCO,  
% commentary by Ale, Ryan, Larry in Stat Sci, a bunch of model X stuff including
% Lukas Janson's recent paper, Eugene + Aaditya's recent paper?).  Larry has put
% some references in the bib.  Put more if we can find them, especially on time
% series.

@@

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable but
regular degree of revision or ``backfill'' after their initial publication date.
The survey-based signal, CTIS-CLI-in-community, is 2 days latent, and rarely
undergoes any revision at all.  The target variable itself, reported COVID-19
case rates, is 1 day latent, and exhibits frequent, unpredictable revisions
(sometimes clear anomalies) after initial publication.  Compared to the revision
pattern in the medical claims signals, which are much more systematic in nature,
revisions in case reports can be highly erratic. Big spikes or other anomalies
can occur in the data as reporting backlogs are cleared, changes in case
definitions are made, etc. Groups like JHU CSSE then work tirelessly to correct
such anomalies after first publication (e.g., they will attempt to
back-distribute a spike when a reporting backlog is cleared, by working with a
local authority to figure out how this should best be done), which can result
in very nontrivial and often unpredictable revisions.

TODO For a figure, use Jacob's ``famous'' example of backfill in DV-CLI?   
\url{https://cmu-delphi.github.io/covidcast/talks/ml-summit/talk.html#(20)}

@@

In addition to the above ``pure'' autoregression model, we consider five
probabilistic forecasting models of the form: 
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary indicators as we do for the case
rates. Training then proceeds just as before: we use the same 7 quantile levels
in \eqref{eq:quantile_levels}, and fit quantile regression separately for each 
level $\tau$, using data from all 306 HRRs and a trailing window of 21 days of
training data.

@@

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j}.  
\end{equation}
In addition to this ``pure'' autoregression model, we also consider five
logistic models of the form: 
\begin{multline}
\label{eq:hotspot_ar_x}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

As described previously, the auxiliary indicators defined in terms of medical
claims, CHNG-CLI, CHNG-COVID, and DV-CLI, undergo a significant and systematic
pattern of revision, or ``backfill'', after their initial publication.  Given
their somewhat statistically-regular backfill profiles, it would be reasonable
to attempt to estimate their finalized values based on vintage data---a problem
we refer to as \textit{nowcasting}---as a pre-processing step before using them
as features in the models in \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x}. 
Nowcasting is itself a highly nontrivial modeling problem, and we do not attempt
it in this paper (though it is a topic of ongoing work in our research group).
We do note, however, that this would likely improve the performance of the
models involving claims-based indicators in particular.

@@

Some signals display a significant degree of spatial heterogeneity, by which we
mean the values that they take at different geographic locations are not 
comparable.  This is true of the Google-AA signal (due to the way in which the 
underlying search trends time series is self-normalized, see
\cite{GoogleSymptoms}) and of the claims-based signals (see the discussion in
our companion paper on the API and database).  Such spatial heterogeneity likely
hurts the performance of the predictive models that rely on these signals,
since the models are trained by pooling data from all locations.  In this paper,
we do not attempt to address this issue (it is again a topic of ongoing work in
our research group), and we simply note that location-specific effects (or 
pre-processing to remove spatial bias) would likely improve the performance of
the models involving Google-AA and the claims-based indicators.

@@

>>>>

There is another, more subtle issue in evaluating predictive utility that 
deserves explicit mention, as it will play a key role in our analysis.
Signals computed from surveillance streams will often be subject to  
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and/or processed late. Correlations between such a signal and case
rates calculated ``after the fact'' (i.e., computed retrospectively, using the
finalized values of this signal) will not deliver an honest answer to the    
question of whether this signal would have been useful in real time. Instead,
we build predictive models using only the data that would have been available
\textit{as of} the prediction date, and compare the ensuing predictions in terms 
of accuracy. To do so, we leverage Delphi's \texttt{evalcast} R package 
\cite{EvalcastR}, which plugs into the COVIDcast API's data versioning system,  
and facilitates honest backtesting.

Finally, it is worth noting that examining the importance of additional features
for prediction is a core question in inferential statistics and econometrics,
with work dating back to at least \cite{Granger:1969}. Still today, drawing
rigorous inference based on predictions, without (or with lean) assumptions, is
an active field of research from both the applied and theoretical angles;
see, e.g., \cite{Diebold:2002, McCraken:2007, Diebold:2015, Stokes:2017,
  Lei:2018, Rinaldo:2019, Williamsom:2020, Zhang:2020, Dai:2021, Fryer:2021}.
Our take on this problem is in line with much of this literature; however, in
order to avoid making any explicit assumptions, we do attempt to make explicit
significance statements, and instead, broadly examine the stability of our
conclusions with respect to numerous modes of analysis.

@@

\begin{figure}[tb!]
\centering
\includegraphics[width=\columnwidth]{fig/vintage.pdf}
\caption{TODO}
\label{fig:vintage}
\end{figure}

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable but
regular degree of revision or ``backfill'' after their initial publication date.
The survey-based signal, CTIS-CLI-in-community, is 2 days latent, and rarely
undergoes any revision at all.  The target variable itself, reported COVID-19
case rates, is 1 day latent, and exhibits frequent, unpredictable revisions
(sometimes clear anomalies) after initial publication.  Compared to the revision
pattern in the medical claims signals, which are much more systematic in nature,
revisions in case reports can be highly erratic. Big spikes or other anomalies
can occur in the data as reporting backlogs are cleared, changes in case
definitions are made, etc. Groups like JHU CSSE then work tirelessly to correct
such anomalies after first publication (e.g., they will attempt to
back-distribute a spike when a reporting backlog is cleared, by working with a
local authority to figure out how this should best be done), which can result
in very nontrivial and often unpredictable revisions.  See Figure
\ref{fig:vintage} for an example.

@@

In addition to the above ``pure'' autoregressive model, we consider five
probabilistic forecasting models of the form: 
\begin{multline}
\label{eq:forecast_ar_x}
\Quantile_\tau(Y_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y_{\ell,t-7j} + 
\sum_{j=0}^2 \gamma^{a,\tau}_j X_{\ell,t-7j},
\end{multline}
where $X_{\ell,t}$ denotes any one of the five auxiliary indicators---CHNG-CLI, 
CHNG-COVID, CTIS-CLI-in-community, DV-CLI, or Google-AA---at location $\ell$ and
time $t$. Note that we apply the same lags (current value, along with the values
7 and 14 days ago) for the auxiliary indicators as we do for the case
rates. Training then proceeds just as before: we use the same 7 quantile levels
in \eqref{eq:quantile_levels}, and fit quantile regression separately for each 
level $\tau$, using data from all 306 HRRs and a trailing window of 21 days of
training data.

@@

We treat hotspot prediction as a binary classification problem and use a setup 
altogether quite similar to the forecasting setup described previously.
Separately for each $a=7,\ldots,28$, to predict $Z_{\ell,t+a}$, we consider a
simple logistic model:
\begin{equation}
\label{eq:hotspot_ar}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, s \leq t)  
= \alpha^{a,\tau} + \sum_{j=0}^2 \^{a,\tau}_j Y^\Delta_{\ell,t-7j}.  
\end{equation}
In addition to this ``pure'' autoregressive model, we also consider five
logistic models of the form: 
\begin{multline}
\label{eq:hotspot_ar_x}
\logit(Z_{\ell,t+a} \,|\, Y_{\ell,s}, X_{\ell,s}, s \leq t)  
= \\ \alpha^{a,\tau} + \sum_{j=0}^2 \beta^{a,\tau}_j Y^\Delta_{\ell,t-7j} +  
\sum_{j=0}^2 \gamma^{a,\tau}_j X^\Delta_{\ell,t-7j},
\end{multline}
where we use \smash{$X^\Delta_{\ell,t} = (X_{\ell,t} - X_{\ell,
    t-7})/(X_{\ell,t-7})$}, and again $X_{\ell,t}$ stands for any of the five  
auxiliary indicators at location $\ell$ and time $t$.  We fit all models
\eqref{eq:hotspot_ar}, \eqref{eq:hotspot_ar_x} using logistic regression, 
pooling all 306 HRRs and using a 21-day trailing window for the training data.

@@

As described previously, the auxiliary indicators defined in terms of medical
claims, CHNG-CLI, CHNG-COVID, and DV-CLI, undergo a significant and systematic
pattern of revision, or ``backfill'', after their initial publication.  Given
their somewhat statistically-regular backfill profiles, it would be reasonable
to attempt to estimate their finalized values based on vintage data---a problem
we refer to as \textit{nowcasting}---as a pre-processing step before using them
as features in the models in \eqref{eq:forecast_ar_x}, \eqref{eq:hotspot_ar_x}. 
Nowcasting is itself a highly nontrivial modeling problem, and we do not attempt
it in this paper (it is a topic of ongoing work in our research group), but we
note that nowcasting would likely improve the performance of the models
involving claims-based signals in particular.

@@

Some signals display a significant degree of spatial heterogeneity, by which we
mean the values that they take at different geographic locations are not 
comparable.  This is true of the Google-AA signal (due to the way in which the 
underlying search trends time series is self-normalized, see
\cite{GoogleSymptoms}) and of the claims-based signals (see the discussion in
our companion paper on the API and database).  Such spatial heterogeneity likely
hurts the performance of the predictive models that rely on these signals,
as we train the models on data pooled over all locations.  In the current 
paper, we do not attempt to address this issue (it is again a topic of ongoing
work in our group), and we simply note that location-specific effects
(or pre-processing to remove spatial bias) would likely improve the performance 
of the models involving Google-AA and the claims-based indicators.

@@

We begin with a summary of the high-level conclusions.  Here, and in the
following subsections, we will use ``AR'' to refer to the pure autoregressive
model both in forecasting \eqref{eq:forecast_ar} and in hotspot prediction
\eqref{eq:hotspot_ar} (where the reference to the task should be clear from the
context). We will also use the name of an auxiliary indicator---``CHNG-CLI'', 
``CHNG-CLI'', ``CHNG-COVID'', ``CTIS-CLI-in-community'', ``DV-CLI'', or
``Google-AA''---interchangeably with the model in forecasting
\eqref{eq:forecast_ar_x} or hotspot prediction \eqref{eq:hotspot_ar_x} that uses
this particular indicator as a feature (the meaning should be clear from the
context).  So, e.g., the CHNG-CLI model in forecasting is the one in
\eqref{eq:forecast_ar_x} that sets $X_{\ell,t}$ to be the value of the CHNG-CLI
indicator at location $\ell$ and time $t$, fits the coefficients based on the
training data, and makes its predictions by setting $X_{\ell,t}$ to the
appropriate CHNG-CLI values.  Finally, we use the term ``indicator model'' to 
broadly refer to any one of the ten models of the form \eqref{eq:forecast_ar_x},
\eqref{eq:hotspot_ar_x} (five in each of the forecasting and hotspot prediction
tasks).

\begin{itemize}
\item Stratifying predictions by the ahead value ($a=7,\ldots,21$), and
  aggregating results over the prediction period (early June through end of
  December 2020), each of the indicator models generally gives a boost in
  predictive accuracy over the AR model, in both the forecasting and hotspot 
  prediction tasks.  The gains in accuracy generally attenuate as the ahead 
  value grows.

\item In the same aggregate view, CHNG-COVID and DV-CLI offer the biggest 
  gains in each of forecasting and hotspot prediction.  CHNG-CLI is
  inconsistent: it provides a big gain in hotspot prediction, but little gain 
  in forecasting (it seems to be hurt by a notable lack of robustness, due
  to backfill).  CTIS-CLI-in-community and Google-AA each provide decent 
  gains in forecasting and hotspot prediction.  The former's performance in 
  forecasting is notable in that it clearly improves on AR even at the largest
  ahead values.

\item In a more detailed analysis of forecasting performance, the indicator
  models tend to be worse than AR when case rates are increasing (this is most
  notable in CHNG-CLI and DV-CLI), and better than AR when case rates are flat
  or decreasing (most notable in CHNG-COV and CTIS-CLI-in-community).
  More rarely does an indicator model tend to beat AR when case rates are
  increasing, but there appears to be some evidence of this for Google-AA. 
  % In an increasing period, when an indicator model is worse than AR, it tends
  % to underpredict relative to AR (its median forecast is lower).  In a flat
  % or decreasing  period, when an indicator model is better than AR, AR tends
  % to overpredict relative to it (the median AR forecast is higher).

\item In this same analysis, when an indicator model performs better than AR  
  in a decreasing period, this tends to co-occur with instances in which the
  indicator ``leads'' case rates (meaning, roughly, on a short-time scale in a
  given location, its behavior mimics that of case rates some number of days
   ahead).  On the other hand, neither leading nor lagging behavior tend to
  co-occur as consistently with an indicator model doing better in periods of
  increase, or worse in periods of increase or decrease.  
\end{itemize}

Finally, to quantify the importance of training and making predictions using
proper vintage data, we ran a parallel set of forecasting and hotspot prediction 
experiments using finalized data. The results, given in the supplement, show
that training and making predictions on finalized data can result in grossly
optimistic estimates of true test-time performance (up to 10\% better in terms
of average WIS or AUC).
